Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-06 paper-new/discussion.tex
several fixes to paper
[paper] add outline of remainder of paper to intro
[paper] update discussion of mechanization

\section{Discussion}

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper,
we are mechanizing our results and developing a reusable framework
for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.
%
As of this writing, the work is in progress, but we have
constructed most parts of the concrete model discussed in
Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations,
and we used the guarded features to define the guarded lift + error monad
and the dynamic type.

We plan to formalize the construction of perturbations and 
quasi-representable relations, but we have yet to decide
whether to follow the approach we take in this work and define
the abstract notion of intensional model and formalize the constructions in that setting,
and then apply those abstract constructions to the concrete model.
Alternatively, it may be better from a mechanization standpoint
to carry out those abstract constructions explicitly in the concrete model,
i.e., our representation of objects in the concrete model of predomains
would include a field for the perturbations and our notion of relations
would include fields for the push-pull property and quasi-representability.
We leave this investigation to future work.

Lastly, we plan to formalize the adequacy result discussed in
Section \ref{sec:adequacy}. This will involve adding axioms about
clock quantification as well as about the clock-irrelevance of
booleans and natural numbers, since as of this writing these
axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
formalized some basic results involving clocks as part of our development.

@@

Working internally to guarded type theory reduces the overhead of
needing to carry around the step-indices in the proofs as is required
when using explicit step-indexing. Additionally, the logical relations
constructed to prove graduality in prior work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
from technical complications of requiring two separate expression
relations, one that counts steps on the left and the other on the
right, and there is no analogue of this in our approach. However,
using two expression relations allows some but not all transitivite
reasoning of term precision to be recovered. In the future we aim to
explore if this approach is feasible in guarded semantics.

>>>>

\section{Discussion}\label{sec:discussion}

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We have formalized the key aspects in the definition of the concrete model
discussed in Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations, and we
used the guarded features to define the guarded lift + error monad and the
dynamic type.

Building on these definitions, we formalized the notion of perturbations and
push-pull structures as well as quasi-representable relations, culminating in
the definition of value and computation types as predomains (resp. error
domains) equipped with a monoid of perturbations with an interpretation
homomorphism into the endomorphisms bisimilar to the identity. We constructed
the predomain for the dynamic type via mixed induction and guarded recursion,
defined the three relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
quasi-representable.

% We plan to formalize the construction of perturbations and 
% quasi-representable relations, but we have yet to decide
% whether to follow the approach we take in this work and define
% the abstract notion of intensional model and formalize the constructions in that setting,
% and then apply those abstract constructions to the concrete model.
% Alternatively, it may be better from a mechanization standpoint
% to carry out those abstract constructions explicitly in the concrete model,
% i.e., our representation of objects in the concrete model of predomains
% would include a field for the perturbations and our notion of relations
% would include fields for the push-pull property and quasi-representability.
% We leave this investigation to future work.

Lastly, we have formalized the adequacy result discussed in Section
\ref{sec:adequacy}. This required us to add axioms about clock quantification as
well as axioms asserting the clock-irrelevance of booleans and natural numbers,
since as of this writing these axioms are not built-in to Guarded Cubical Agda.
The mechanization of adequacy also required us to formalize some essential
lemmas involving clocks and clock-irrelevance; we may later refactor these as
part of a ``standard library'' for Guarded Cubical Agda.

@@

Working internally to guarded type theory reduces the overhead of
needing to carry around the step-indices in the proofs as is required
when using explicit step-indexing. Additionally, the logical relations
constructed to prove graduality in prior work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
from technical complications of requiring two separate expression
relations, one that counts steps on the left and the other on the
right, and there is no analogue of this in our approach. However,
using two expression relations allows some but not all transitive
reasoning of term precision to be recovered. In the future we aim to
explore if this approach is feasible in guarded semantics.
